---
title: "Within- and between-cluster correlations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Within- and between-cluster correlations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 6
)
```

Repeated observations contain at least two kinds of association. A
between-cluster correlation asks whether clusters with higher values on one
variable also tend to have higher values on another. A within-cluster
correlation asks whether deviations from a cluster's usual level move together.
These correlations can differ in size or even have opposite signs, so a single
correlation calculated from all rows can be misleading.

This vignette demonstrates the basic `wbCorr` workflow with repeated
measurements nested in people.

## A contrasting example

The data below are constructed so that people with higher average `x` also
have higher average `y`. Within a person, however, occasions with higher `x`
than usual tend to have lower `y` than usual.

```{r simulate-data}
set.seed(2026)

n_people <- 60
n_occasions <- 10
person <- rep(seq_len(n_people), each = n_occasions)

person_x <- rnorm(n_people, sd = 2)
person_y <- 0.8 * person_x + rnorm(n_people, sd = 0.5)
occasion_x <- rnorm(n_people * n_occasions)
occasion_y <- -0.7 * occasion_x + rnorm(n_people * n_occasions, sd = 0.6)

dat <- data.frame(
  person = person,
  x = rep(person_x, each = n_occasions) + occasion_x,
  y = rep(person_y, each = n_occasions) + occasion_y
)

round(cor(dat$x, dat$y), 2)
```

The pooled correlation is positive, but it combines the two levels of
association.

## Separate the two levels

Pass the data and the name of the cluster variable to `wbCorr()`. Using
`inference = "none"` is convenient when only the coefficients are needed.

```{r fit}
library(wbCorr)

fit <- wbCorr(
  data = dat,
  cluster = "person",
  inference = "none"
)

fit
```

`get_table()` returns the full within- and between-cluster result tables.

```{r tables}
tables <- get_table(fit)

tables$within
tables$between
```

Here the within-person correlation is negative while the between-person
correlation is positive. The distinction answers two different questions; one
coefficient should not be substituted for the other.

## Correlation matrices and plots

Use `summary()` or `get_matrix()` to retrieve matrices. The merged matrix places
within-cluster correlations above the diagonal and between-cluster correlations
below it.

```{r matrices}
matrices <- summary(fit)
matrices$merged_wb
```

The corresponding centered data can also be inspected visually.

```{r within-plot, fig.cap = "Within-person association after subtracting each person's variable means."}
plot(fit, "within")
```

```{r between-plot, fig.cap = "Between-person association among person-level variable means."}
plot(fit, "between")
```

## Add uncertainty when needed

By default, Pearson correlations include analytic tests and confidence
intervals. These are working approximations for clustered observations. For
intervals that preserve top-level dependence, resample whole clusters instead.

```{r bootstrap, eval = FALSE}
set.seed(2026)

fit_boot <- wbCorr(
  data = dat,
  cluster = "person",
  inference = "cluster_bootstrap",
  nboot = 1000
)
```

Use substantially more than the technical minimum number of bootstrap
replicates for substantive analyses, and inspect `n_boot_valid` in the detailed
tables. See `?wbCorr` for weighting, centering, missing-data, Spearman, and
diagnostic options.
