---
title: "Within- and between-cluster correlations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Within- and between-cluster correlations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 6
)
```

Repeated observations contain at least two kinds of association. A
between-cluster correlation asks whether clusters with higher values on one
variable also tend to have higher values on another. A within-cluster
correlation asks whether deviations from a cluster's usual level move together.
These correlations can differ in size or even have opposite signs, so a single
correlation calculated from all rows can be misleading.

This vignette demonstrates the basic `wbCorr` workflow with repeated
measurements nested in people.

## A contrasting example

The data below are constructed so that people with higher average `x` also
have higher average `y`. Within a person, however, occasions with higher `x`
than usual tend to have lower `y` than usual.

```{r simulate-data}
set.seed(2026)

n_people <- 60
n_occasions <- 10
person <- rep(seq_len(n_people), each = n_occasions)

person_x <- rnorm(n_people, sd = 2)
person_y <- 0.8 * person_x + rnorm(n_people, sd = 0.5)
occasion_x <- rnorm(n_people * n_occasions)
occasion_y <- -0.7 * occasion_x + rnorm(n_people * n_occasions, sd = 0.6)

dat <- data.frame(
  person = person,
  x = rep(person_x, each = n_occasions) + occasion_x,
  y = rep(person_y, each = n_occasions) + occasion_y
)

round(cor(dat$x, dat$y), 2)
```

The pooled correlation is positive, but it combines the two levels of
association.

## Separate the two levels

Pass the data and the name of the cluster variable to `wbCorr()`. Using
`inference = "none"` is convenient when only the coefficients are needed.

```{r fit}
library(wbCorr)

fit <- wbCorr(
  data = dat,
  cluster = "person",
  inference = "none"
)
```

## Inspect the correlation matrices

For interactive use, `summary()` provides short selectors for the most useful
correlation matrices. Request the within-cluster matrix with `"w"` and the
between-cluster matrix with `"b"`.

```{r summarize-levels}
summary(fit, "w")
summary(fit, "b")
```

Here the within-person correlation is negative while the between-person
correlation is positive. The distinction answers two different questions; one
coefficient should not be substituted for the other.

For a compact combined display, use `"wb"`. Within-cluster correlations appear
above the diagonal, between-cluster correlations appear below it, and ICCs
appear on the diagonal. Use `"bw"` to reverse the two triangles.

```{r summarize-merged}
summary(fit, "wb")
```

Calling `print(fit)` prints a compact preview of the detailed result tables.

## Plot the two levels

The corresponding centered data can be inspected visually.

```{r within-plot, fig.cap = "Within-person association after subtracting each person's variable means."}
plot(fit, "within")
```

```{r between-plot, fig.cap = "Between-person association among person-level variable means."}
plot(fit, "between")
```

## Extract results for further work

The `summary()` shortcuts are convenient for inspection. For programmatic use,
`get_table()` returns the full results and pair-specific diagnostics, while
`get_matrix()` returns selected matrices. Set `numeric = TRUE` when unrounded
numeric coefficients are needed for downstream calculations.

```{r extract-results}
tables <- get_table(fit)
within_table <- tables$within

within_matrix <- get_matrix(fit, "w", numeric = TRUE)$within
merged_matrix <- get_matrix(fit, "wb", numeric = TRUE)$merged_wb
```

## Add a third nested level

For observations nested in lower-level units nested in upper-level units, pass
a fully named list ordered lower-to-higher. The names describe the grouping
levels; the values name columns in the data. The example below has occasions
within people within dyads.

```{r simulate-three-level}
set.seed(2027)

n_dyads <- 30
n_people_per_dyad <- 2
n_occasions_per_person <- 8
n_people <- n_dyads * n_people_per_dyad
n_rows <- n_people * n_occasions_per_person

dyad <- rep(seq_len(n_dyads),
            each = n_people_per_dyad * n_occasions_per_person)
person <- rep(seq_len(n_people), each = n_occasions_per_person)

dyad_x <- rnorm(n_dyads)
dyad_y <- 0.7 * dyad_x + rnorm(n_dyads, sd = 0.7)
person_x <- rnorm(n_people)
person_y <- -0.5 * person_x + rnorm(n_people, sd = 0.8)
occasion_x <- rnorm(n_rows)
occasion_y <- 0.4 * occasion_x + rnorm(n_rows)

nested_dat <- data.frame(
  dyad = dyad,
  person = person,
  x = rep(dyad_x, each = n_people_per_dyad * n_occasions_per_person) +
    rep(person_x, each = n_occasions_per_person) + occasion_x,
  y = rep(dyad_y, each = n_people_per_dyad * n_occasions_per_person) +
    rep(person_y, each = n_occasions_per_person) + occasion_y
)
```

```{r fit-three-level}
nested_fit <- wbCorr(
  data = nested_dat,
  cluster = list(person = "person", dyad = "dyad"),
  inference = "none"
)

summary(nested_fit, "l1")
summary(nested_fit, "l2")
summary(nested_fit, "l3")
```

`level1` correlates occasion deviations from person means. `level2` correlates
person means after subtracting the corresponding dyad mean. `level3`
correlates dyad means. All quantities are computed separately for each
variable pair: observations contribute equally at level 1, people contribute
equally at level 2 and to each dyad mean, and dyads contribute equally at
level 3.

With unequal numbers of observations per person, this equal-person dyad mean
is deliberately different from the observation-weighted upper mean used in
conventional manifest contextual-model decompositions
([Kerkhoff & Nussbeck, 2023](https://doi.org/10.5964/meth.9775)). The two
coincide when people contribute the same number of observations. These are
descriptive correlations, not estimates of a multilevel regression, R-squared,
or latent parameter.

The three matrices answer three different questions and are therefore not
combined into a two-triangle display. Likewise, a single one-way ICC is not
reported for a three-level object.

```{r extract-three-level}
nested_tables <- get_table(nested_fit)
nested_matrices <- get_matrix(nested_fit, numeric = TRUE)

nested_tables$level2
nested_matrices$level3
```

The first implementation requires globally unique lower-level IDs, strict
nesting, Pearson correlations, pairwise-complete centering, and equal-unit
weighting. Use `missing_data = "listwise"` when all matrices should use one
common row sample.

## Add uncertainty when needed

By default, Pearson correlations include analytic tests and confidence
intervals. These are working approximations for clustered observations. For
intervals that preserve top-level dependence, resample whole clusters instead.

```{r bootstrap, eval = FALSE}
set.seed(2026)

fit_boot <- wbCorr(
  data = dat,
  cluster = "person",
  inference = "cluster_bootstrap",
  nboot = 1000
)
```

Use substantially more than the technical minimum number of bootstrap
replicates for substantive analyses, and inspect `n_boot_valid` in the detailed
tables. See `?wbCorr` for weighting, centering, missing-data, Spearman, and
diagnostic options.

For a three-level fit, the same inference selector resamples intact upper-level
units and preserves all of their lower units and observations:

```{r bootstrap-three-level, eval = FALSE}
set.seed(2027)

nested_boot <- wbCorr(
  data = nested_dat,
  cluster = list(person = "person", dyad = "dyad"),
  inference = "cluster_bootstrap",
  nboot = 1000
)
```

Analytic three-level inference is not supplied because its operating
characteristics have not been justified for these descriptive estimands.
